Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-04-06 elsevier_MANUSCRIPT.qmd
prepping for submission

-   Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
-   A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse low-flow streamgage network, typical of drylands.
-   Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.

@@

The study area represents a range of heterogeneous climate, elevation,

The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$. Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south. The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim. The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft). Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation. The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft). Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually [@az_climate2024].

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. While winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]. Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

@@

![Trends in BFI over full period of record for instrumented sites used in this study. Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%. White circles represent sites with no statistically significant trends.](images/updated/BFITrends_20250404.png){#fig-instrumented-trend}

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN). The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly base flow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.

>>>>

-   Approximately 32% of Arizona's streamflow originates from groundwater discharge.
-   XGBoost predicted long-term BFI in ungauged basins, filling dryland data gaps.
-   Baseflow trends track precipitation, emphasizing climate’s role in dryland hydrology.

@@

Arizona, located in the southwestern United States, spans approximately 295,253 km² and encompasses a diverse range of landscapes, elevations, and climate regimes. The state includes portions of three major physiographic regions: the Colorado Plateau in the northeast, the Basin and Range province in the south and west, and the Central Highlands, a transitional zone between them. This heterogeneity results in substantial variation in environmental conditions across the state. The Colorado Plateau is characterized by high-elevation desert and mountain woodlands, averaging 1,936 masl (6,352 ft), with mean temperatures ranging from -6°C (20°F) to 26°C (80°F) and annual precipitation of about 580 mm (23 in). In contrast, the Basin and Range region is lower in elevation, averaging 490 masl (1,600 ft), and features a semi-arid to arid climate, with temperatures ranging from 15°C (60°F) to 43°C (110°F) and an average annual precipitation of 200 mm (8 in) [@az_climate2024]. The Central Highlands feature a mix of mountainous terrain and interspersed basins, adding to the state's topographic and hydroclimatic complexity.

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In summer, localized and intense convective storms stem from the North American Monsoon; in winter, orographic precipitation is delivered by Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. Though winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]. Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages reduces potential errors [@reitz2017].

@@

![Trends in BFI over full period of record for instrumented sites used in this study. Red downward (blue upward) arrows indicate a decreasing (increasing) trend at a significance level of 5%. White circles represent sites with no statistically significant trends.](images/updated/BFITrends_20250404.png){#fig-instrumented-trend}

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an ANN. The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly base flow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.
